---
title: "Bayesian Logistic Regression for Heart Disease Prediction"
date: "2025-12-05"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rjags)
library(caret)
library(loo)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ROCR)
```


```{r}
df <- read.csv("heart-disease.csv")
head(df)
```

```{r}
str(df)
```


```{r}
# Check Distribution of Target Column (target variable for classification)
table(df$target)
```

```{r}
# The proportion of the positive class (class 1) in target column is used as the probability threshold for class prediction
P1 = (165/(165+138))
P1
```

```{r}
#Split Dataset -> Train (80%) and test (20%)
set.seed(123)
idx <- createDataPartition(df$thal, p = 0.8, list = FALSE)

train <- df[idx,]
test <- df[-idx,]
```

```{r}
robust_params <- function(x){
  list(
    med = median(x, na.rm = TRUE),
    iqr = IQR(x, na.rm = TRUE)
  )
}

num_cols <- c("age", "trestbps", "chol", "thalach", "oldpeak")
params <- lapply(train[num_cols], robust_params)

robust_scale_with_params <- function(x, p){
  (x - p$med) / p$iqr
}

for(col in num_cols){
  train[[col]] <- robust_scale_with_params(train[[col]], params[[col]])
}

for(col in num_cols){
  test[[col]] <- robust_scale_with_params(test[[col]], params[[col]])
}


x_train <- model.matrix(~ ., data = train[-14])
x_test  <- model.matrix(~ ., data = test[-14])

y_train <- train$target
y_test  <- test$target

n <- nrow(x_train)
p <- ncol(x_train)

```

## Bayesian Logistic Regression Model 1
```{r}
# Model Bayesian Logistic Regression 1
bayes_model1 <- textConnection("model{ 
            # Likelihood
            for(i in 1:n){ 
                  Y[i] ~ dbern(pi[i]) 
                  logit(pi[i]) <- inprod(beta[], X[i, ])
                  
                  # Posterior Predictive Samples
                  Y2[i] ~ dbern(pi[i])
                  
                  log_lik[i] <- logdensity.bern(Y[i], pi[i])
            } 
            
            
            # Priors for Regression Coefficients 
            for(j in 1:p){
                beta[j] ~ dnorm(0,0.5)
            } 
            
}")

data1 <- list(Y = y_train, X = x_train, n = n, p = p)
burn1 <- 10000
iter1 <- 50000

model1 <- jags.model(bayes_model1, data = data1, n.chains = 3)
update(model1, burn1)
samp1 <- coda.samples(model1, variable.names = c("beta", "pi", "Y2", "log_lik"), n.iter = iter1, n.thin = 5)
```

```{r}
beta_samp <- samp1[, grep("beta", colnames(samp1[[1]]))]
summary(beta_samp)
```

## Visualization
```{r}
var_names <- c(
  "Intercept",
  colnames(x_train)[-1]
)
mb <- as.matrix(beta_samp)

for (k in seq_along(colnames(mb))) {

  param <- colnames(mb)[k]        
  label <- var_names[k]          
  
  df <- data.frame(value = mb[, k])
  
  # 95% credible interval
  ci <- quantile(df$value, probs = c(0.025, 0.975))
  lower <- ci[1]
  upper <- ci[2]


  # density estimation
  dens <- density(df$value)
  dens_df <- data.frame(x = dens$x, y = dens$y)
  
  # CI shading region only
  dens_ci <- subset(dens_df, x >= lower & x <= upper)

  p <- ggplot() +
    # full density curve outline
    geom_line(data = dens_df, aes(x, y), color = "black", size = 1) +
    # shaded CI only
    geom_area(data = dens_ci, aes(x, y), fill = "skyblue", alpha = 0.4) +
    # posterior mean (red dashed)
    geom_vline(xintercept = mean(df$value),
               linetype = "dashed", color = "red") +
    
    labs(
      title = paste0("Posterior Density of ", label, " (", param, ")"),
      subtitle = paste0("95% CI: [", round(lower,3), ", ", round(upper,3), "]"),
      x = param, y = "Density"
    ) +
    theme_minimal()
  
  print(p)
}

```

```{r}
par(mar = c(4, 4, 2, 2))
plot(beta_samp)
```


```{r}
# Gelman-Rubin diagnostic (multivariate psrf > 1.1 suggest poor convergence)
gelman.diag(beta_samp)
```
All PSRF (Potential Scale Reduction Factor) values for the parameters are below 1.1, and the multivariate PSRF is 1, which is also under the commonly used threshold of 1.1. These results indicate that the MCMC chains have likely converged to the target posterior distribution.

## Model Evaluation
### Posterior Predictive Check
```{r}
m_ppc <- do.call(rbind, lapply(samp1, as.matrix))

beta_cols <- grep("^beta\\[", colnames(m_ppc))
y2_cols   <- grep("^Y2\\[",   colnames(m_ppc))

beta_mat  <- m_ppc[, beta_cols, drop = FALSE]
y2_mat    <- m_ppc[, y2_cols,   drop = FALSE]

logits_post <- x_train %*% t(beta_mat)
p_post <- plogis(logits_post)
p_mean <- rowMeans(p_post)

plot_data <- data.frame(
  Observed = y_train,
  Prob     = as.numeric(p_mean)
)

plot_data$Bin <- cut(
  plot_data$Prob,
  breaks = quantile(plot_data$Prob, probs = seq(0, 1, 0.1)),
  include.lowest = TRUE,
  labels = FALSE
)

obs_props <- plot_data %>%
  group_by(Bin) %>%
  summarise(Proportion = mean(Observed), .groups = "drop") %>%
  mutate(Type = "Observed")

sample_indices <- 1:nrow(y2_mat)
n_draws <- length(sample_indices)

ppc_list <- vector("list", n_draws)

for(i in sample_indices) {

  y_sim <- y2_mat[i, ]

  tmp_df <- data.frame(
    Bin = plot_data$Bin,
    Simulated = y_sim
  )

  sim_summary <- tmp_df %>%
    group_by(Bin) %>%
    summarise(Proportion = mean(Simulated), .groups = "drop") %>%
    mutate(Iter = i)

  ppc_list[[i]] <- sim_summary
}

ppc_data <- do.call(rbind, ppc_list)

ppc_mean <- ppc_data %>%
  group_by(Bin) %>%
  summarise(Proportion = mean(Proportion), .groups = 'drop')

ggplot() +
  geom_line(data = ppc_data,
            aes(x = Bin, y = Proportion, group = Iter),
            color = "blue", alpha = 0.01) +

  geom_line(data = ppc_mean,
            aes(x = Bin, y = Proportion),
            color = "orange", linetype = "dashed", size = 1.2) +
  geom_point(data = ppc_mean,
             aes(x = Bin, y = Proportion),
             color = "orange", shape = 17, size = 3) +

  geom_line(data = obs_props,
            aes(x = Bin, y = Proportion),
            color = "red", size = 1.2) +
  geom_point(data = obs_props,
             aes(x = Bin, y = Proportion),
             color = "red", size = 3) +

  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "PPC",
    x = "Bin Index",
    y = "Proportion"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.1, face = "bold", size = 14))
```



### Model Fit Statistic
#### DIC
```{r}
dic1 <- dic.samples(model1, n.iter = iter1, n.thin = 5, progress.bar = "none")
dic1
```

#### LOO
```{r}
m <- as.matrix(samp1)
log_lik_mat <- m[, grep("^log_lik\\[", colnames(m))]
dim(log_lik_mat)
```

```{r}
s <- nrow(log_lik_mat)
chain_id <- rep(1:3, each = s / 3)

r_eff   <- relative_eff(exp(log_lik_mat), chain_id = chain_id)
loo1 <- loo(log_lik_mat, r_eff = r_eff)
loo1
```


#### WAIC
```{r}
waic1 <- waic(log_lik_mat)
waic1
```


```{r}
get_eval_metrics <- function(y_true, y_pred, type) {
  TP <- sum(y_pred == 1 & y_true == 1)
  FP <- sum(y_pred == 1 & y_true == 0)
  TN <- sum(y_pred == 0 & y_true == 0)
  FN <- sum(y_pred == 0 & y_true == 1)
  
  accuracy  <- (TP + TN) / (TP + TN + FP + FN)
  precision <- TP / (TP + FP)
  recall    <- TP / (TP + FN)  
  f1_score  <- 2 * (precision * recall) / (precision + recall)
  
  if(is.nan(precision)) precision <- 0
  if(is.nan(f1_score)) f1_score <- 0
  
  cat(paste0("\n", type, ":\n"))
  cat(paste("Accuracy: ", round(accuracy, 4), "\n"))
  cat(paste("Precision:", round(precision, 4), "\n"))
  cat(paste("Recall:   ", round(recall, 4), "\n"))
  cat(paste("F1 Score: ", round(f1_score, 4), "\n"))
  
  return(list(
    accuracy  = accuracy,
    precision = precision,
    recall    = recall,
    f1_score  = f1_score
  ))
}
```

```{r}
pi_post <- as.matrix(samp1)[ , grep("^pi\\[", colnames(as.matrix(samp1))) ]
p_train_mean <- colMeans(pi_post)
y_pred_train <- ifelse(p_train_mean > 0.5445545, 1, 0) #perhitungan 0.5445545 terdapat di bagian atas (P1 = 0.5445545)

metric_train <- get_eval_metrics(y_true = y_train, y_pred = y_pred_train, "Bayesian Model 1 (Train)")

```

```{r}
beta_mat <- as.matrix(beta_samp)
beta_hat <- colMeans(beta_mat)   

logits_test <- x_test %*% beta_hat
p_test_mean <- 1 / (1 + exp(-logits_test))
y_pred_test <- ifelse(p_test_mean > 0.5445545, 1, 0)

metric_test <- get_eval_metrics(y_true = y_test, y_pred = y_pred_test, "Bayesian Model 1 (Test)")

```


```{r}
m_roc <- do.call(rbind, lapply(samp1, as.matrix))

beta_cols <- grep("^beta\\[", colnames(m_roc))

beta_mat  <- m_roc[, beta_cols, drop = FALSE]

logits_post <- x_test %*% t(beta_mat)
p_post <- plogis(logits_post)
p_mean <- rowMeans(p_post)

bayes_prob1 <- p_mean

pred_ob1 <- prediction(bayes_prob1, y_test)

roc_perf1 <- performance(pred_ob1, measure = "tpr", x.measure = "fpr")

roc_auc_obj1 <- performance(pred_ob1, measure = "auc")
roc_auc_val1 <- unlist(roc_auc_obj1@y.values)

pr_perf1 <- performance(pred_ob1, measure = "prec", x.measure = "rec")

recall_vals1 <- pr_perf1@x.values[[1]]
precision_vals1 <- pr_perf1@y.values[[1]]

precision_vals1[is.nan(precision_vals1)] <- 1

id1 <- order(recall_vals1)
pr_auc_val1 <- sum(
  diff(recall_vals1[id1]) *
    (precision_vals1[id1][-1] + precision_vals1[id1][-length(precision_vals1)]) / 2
)

plot(roc_perf1, col = "blue", lwd = 2, main = "")
title(
  main = paste("Bayesian ROC Curve 1 (AUC =", round(roc_auc_val1, 2), ")"),
  line = 2
)

plot(pr_perf1, col = "blue", lwd = 2, main = "", ylim = c(0, 1))
title(
  main = paste("Bayesian PR Curve 1 (AUC =", round(pr_auc_val1, 2), ")"),
  line = 2
)
```

## Bayesian Logistic Regression Model 2
```{r}
x_train <- model.matrix(~ ., data = train[-14])
x_test  <- model.matrix(~ ., data = test[-14])

y_train <- train$target
y_test  <- test$target

n <- nrow(x_train)
p <- ncol(x_train)
```


```{r}
bayes_model2 <- textConnection("model{ 
            # Likelihood
            for(i in 1:n){ 
                  Y[i] ~ dbern(pi[i]) 
                  logit(pi[i]) <- inprod(beta[], X[i, ])
                  
                  # Posterior Predictive Samples
                  Y2[i] ~ dbern(pi[i])
                  
                  log_lik[i] <- logdensity.bern(Y[i], pi[i])
            } 
            
            
            # Priors for Regression Coefficients 
            for(j in 1:p){
                beta[j] ~ dnorm(0,1)
            } 
            
}")

data2 <- list(Y = y_train, X = x_train, n = n, p = p)
burn2 <- 10000
iter2 <- 50000

model2 <- jags.model(bayes_model2, data = data2, n.chains = 3)
update(model2, burn2)
samp2 <- coda.samples(model2, variable.names = c("beta", "pi", "Y2", "log_lik"), n.iter = iter2, n.thin = 5)
```

```{r}
beta_samp2 <- samp2[, grep("beta", colnames(samp2[[1]]))]
summary(beta_samp2)
```

```{r}
var_names <- c(
  "Intercept",
  colnames(x_train)[-1]
)
mb2 <- as.matrix(beta_samp2)

for (k in seq_along(colnames(mb2))) {

  param <- colnames(mb2)[k]        
  label <- var_names[k]          
  
  df <- data.frame(value = mb2[, k])
  
  # 95% credible interval
  ci <- quantile(df$value, probs = c(0.025, 0.975))
  lower <- ci[1]
  upper <- ci[2]


  # density estimation
  dens <- density(df$value)
  dens_df <- data.frame(x = dens$x, y = dens$y)
  
  # CI shading region only
  dens_ci <- subset(dens_df, x >= lower & x <= upper)

  p <- ggplot() +
    # full density curve outline
    geom_line(data = dens_df, aes(x, y), color = "black", size = 1) +
    # shaded CI only
    geom_area(data = dens_ci, aes(x, y), fill = "skyblue", alpha = 0.4) +
    # posterior mean (red dashed)
    geom_vline(xintercept = mean(df$value),
               linetype = "dashed", color = "red") +
    
    labs(
      title = paste0("Posterior Density of ", label, " (", param, ")"),
      subtitle = paste0("95% CI: [", round(lower,3), ", ", round(upper,3), "]"),
      x = param, y = "Density"
    ) +
    theme_minimal()
  
  print(p)
}
```



```{r}
par(mar = c(4, 4, 2, 2))
plot(beta_samp2)
```

```{r}
# Gelman-Rubin diagnostic (multivariate psrf > 1.1 suggest poor convergence)
gelman.diag(beta_samp2)
```
All PSRF (Potential Scale Reduction Factor) values for the parameters are below 1.1, and the multivariate PSRF is 1, which is also under the commonly used threshold of 1.1. These results indicate that the MCMC chains have likely converged to the target posterior distribution.

## Model Evaluation
### Posterior Predictive Check
```{r}
m_ppc2 <- do.call(rbind, lapply(samp2, as.matrix))

beta_cols2 <- grep("^beta\\[", colnames(m_ppc2))
y2_cols2 <- grep("^Y2\\[",   colnames(m_ppc2))

beta_mat2  <- m_ppc2[, beta_cols2, drop = FALSE]
y2_mat2  <- m_ppc2[, y2_cols2,   drop = FALSE]

logits_post2 <- x_train %*% t(beta_mat2)
p_post2 <- plogis(logits_post2)
p_mean2 <- rowMeans(p_post2)

plot_data2 <- data.frame(
  Observed = y_train,
  Prob     = as.numeric(p_mean2)
)

plot_data2$Bin <- cut(
  plot_data2$Prob,
  breaks = quantile(plot_data2$Prob, probs = seq(0, 1, 0.1)),
  include.lowest = TRUE,
  labels = FALSE
)

obs_props2 <- plot_data2 %>%
  group_by(Bin) %>%
  summarise(Proportion = mean(Observed), .groups = "drop") %>%
  mutate(Type = "Observed")

sample_indices2 <- 1:nrow(y2_mat2)
n_draws2 <- length(sample_indices2)

ppc_list2 <- vector("list", n_draws2)

for(i in sample_indices2) {

  y_sim <- y2_mat2[i, ]

  tmp_df <- data.frame(
    Bin = plot_data2$Bin,
    Simulated = y_sim
  )

  sim_summary <- tmp_df %>%
    group_by(Bin) %>%
    summarise(Proportion = mean(Simulated), .groups = "drop") %>%
    mutate(Iter = i)

  ppc_list2[[i]] <- sim_summary
}

ppc_data2 <- do.call(rbind, ppc_list2)

ppc_mean2 <- ppc_data2 %>%
  group_by(Bin) %>%
  summarise(Proportion = mean(Proportion), .groups = 'drop')

ggplot() +
  geom_line(data = ppc_data2,
            aes(x = Bin, y = Proportion, group = Iter),
            color = "blue", alpha = 0.01) +

  # Mean simulated (orange dashed line)
  geom_line(data = ppc_mean2,
            aes(x = Bin, y = Proportion),
            color = "orange", linetype = "dashed", size = 1.2) +
  geom_point(data = ppc_mean2,
             aes(x = Bin, y = Proportion),
             color = "orange", shape = 17, size = 3) +

  # Observed data (red line)
  geom_line(data = obs_props2,
            aes(x = Bin, y = Proportion),
            color = "red", size = 1.2) +
  geom_point(data = obs_props2,
             aes(x = Bin, y = Proportion),
             color = "red", size = 3) +

  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "PPC",
    x = "Bin Index",
    y = "Proportion"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.1, face = "bold", size = 14))
```

### Model Fit Statistic
#### DIC
```{r}
dic2 <- dic.samples(model2, n.iter = iter2, n.thin = 5, progress.bar = "none")
dic2
```

#### LOO
```{r}
m2 <- as.matrix(samp2)
log_lik_mat2 <- m2[, grep("^log_lik\\[", colnames(m2))]
dim(log_lik_mat2)
```

```{r}
s2 <- nrow(log_lik_mat2)
chain_id2 <- rep(1:3, each = s2 / 3)

r_eff2   <- relative_eff(exp(log_lik_mat2), chain_id = chain_id2)
loo2 <- loo(log_lik_mat2, r_eff = r_eff2)
loo2
```


#### WAIC
```{r}
waic2 <- waic(log_lik_mat2)
waic2
```


```{r}
get_eval_metrics <- function(y_true, y_pred, type) {
  TP <- sum(y_pred == 1 & y_true == 1)
  FP <- sum(y_pred == 1 & y_true == 0)
  TN <- sum(y_pred == 0 & y_true == 0)
  FN <- sum(y_pred == 0 & y_true == 1)
  
  accuracy  <- (TP + TN) / (TP + TN + FP + FN)
  precision <- TP / (TP + FP)
  recall    <- TP / (TP + FN)  
  f1_score  <- 2 * (precision * recall) / (precision + recall)
  
  if(is.nan(precision)) precision <- 0
  if(is.nan(f1_score)) f1_score <- 0
  
  cat(paste0("\n", type, ":\n"))
  cat(paste("Accuracy: ", round(accuracy, 4), "\n"))
  cat(paste("Precision:", round(precision, 4), "\n"))
  cat(paste("Recall:   ", round(recall, 4), "\n"))
  cat(paste("F1 Score: ", round(f1_score, 4), "\n"))
  
  return(list(
    accuracy  = accuracy,
    precision = precision,
    recall    = recall,
    f1_score  = f1_score
  ))
}
```

```{r}
pi_post2 <- as.matrix(samp2)[ , grep("^pi\\[", colnames(as.matrix(samp2))) ]

p_train_mean2 <- colMeans(pi_post2)

y_pred_train2 <- ifelse(p_train_mean2 > 0.5445545, 1, 0)

metric_train2 <- get_eval_metrics(y_true = y_train, y_pred = y_pred_train2, "Bayesian Model 2 (Train)")


```

```{r}
beta_mat2 <- as.matrix(beta_samp2)
beta_hat2 <- colMeans(beta_mat2)   

logits_test2 <- x_test %*% beta_hat2
p_test_mean2 <- 1 / (1 + exp(-logits_test2))

y_pred_test2 <- ifelse(p_test_mean2 > 0.5445545, 1, 0)

metric_test2 <- get_eval_metrics(y_true = y_test, y_pred = y_pred_test2, "Bayesian Model 2 (Test)")

```

```{r}
m_roc2 <- do.call(rbind, lapply(samp2, as.matrix))

beta_cols2 <- grep("^beta\\[", colnames(m_roc2))

beta_mat2 <- m_roc2[, beta_cols2, drop = FALSE]

logits_post2 <- x_test %*% t(beta_mat2)
p_post2 <- plogis(logits_post2)
p_mean2 <- rowMeans(p_post2)

bayes_prob2 <- p_mean2

pred_ob2 <- prediction(bayes_prob2, y_test)

roc_perf2 <- performance(pred_ob2, measure = "tpr", x.measure = "fpr")

roc_auc_obj2 <- performance(pred_ob2, measure = "auc")
roc_auc_val2 <- unlist(roc_auc_obj2@y.values)

pr_perf2 <- performance(pred_ob2, measure = "prec", x.measure = "rec")

recall_vals2 <- pr_perf2@x.values[[1]]
precision_vals2 <- pr_perf2@y.values[[1]]

precision_vals2[is.nan(precision_vals2)] <- 1

id2 <- order(recall_vals2)
pr_auc_val2 <- sum(
  diff(recall_vals2[id2]) *
    (precision_vals2[id2][-1] + precision_vals2[id2][-length(precision_vals2)]) / 2
)

plot(roc_perf2, col = "blue", lwd = 2, main = "")
title(
  main = paste("Bayesian ROC Curve 2 (AUC =", round(roc_auc_val2, 2), ")"),
  line = 2
)

plot(pr_perf2, col = "blue", lwd = 2, main = "", ylim = c(0, 1))
title(
  main = paste("Bayesian PR Curve 2 (AUC =", round(pr_auc_val2, 2), ")"),
  line = 2
)
```


## Frequentist Logistic Regression
```{r}
model_glm <- glm(
  target ~ .,
  data = train,
  family = binomial(link = "logit")
)

summary(model_glm)
```

```{r}
glm_prob <- predict(model_glm, newdata = test, type = "response")
glm_pred <- ifelse(glm_prob > 0.5445545, 1, 0)
metric_glm <- get_eval_metrics(y_test, glm_pred, "Frequentist (test)")
```

```{r}
pred_ob <- prediction(glm_prob, y_test)

roc_perf <- performance(pred_ob, measure = "tpr", x.measure = "fpr")

roc_auc_obj <- performance(pred_ob, measure = "auc")
roc_auc_val <- unlist(roc_auc_obj@y.values)

pr_perf <- performance(pred_ob, measure = "prec", x.measure = "rec")

recall_vals <- pr_perf@x.values[[1]]
precision_vals <- pr_perf@y.values[[1]]

precision_vals[is.nan(precision_vals)] <- 1

id <- order(recall_vals)
pr_auc_val <- sum(
  diff(recall_vals[id]) *
    (precision_vals[id][-1] + precision_vals[id][-length(precision_vals)]) / 2
)

plot(roc_perf, col = "blue", lwd = 2, main = "")
title(
  main = paste("Frequentist ROC Curve (AUC =", round(roc_auc_val, 2), ")"),
  line = 2
)

plot(pr_perf, col = "blue", lwd = 2, main = "", ylim = c(0, 1))
title(
  main = paste("Frequentist PR Curve (AUC =", round(pr_auc_val, 2), ")"),
  line = 2
)
```

```{r}
results_table <- data.frame(
  Model = c("Bayesian Model 1 [Normal(0, 0.5)]",
            "Bayesian Model 2 [Normal(0, 1)]",
            "Frequentist"),
  
  Accuracy  = c(metric_test$accuracy,
                metric_test2$accuracy,
                metric_glm$accuracy),
  
  Precision = c(metric_test$precision,
                metric_test2$precision,
                metric_glm$precision),
  
  Recall    = c(metric_test$recall,
                metric_test2$recall,
                metric_glm$recall),
  
  F1        = c(metric_test$f1_score,
                metric_test2$f1_score,
                metric_glm$f1_score),
  
  ROC       = c(roc_auc_val1, 
                roc_auc_val2, 
                roc_auc_val),
  
  PR        = c(pr_auc_val,
                pr_auc_val1,
                pr_auc_val2)
)

results_table[,-1] <- round(results_table[,-1], 4)
results_table

```

